---
title: "LH Final Project BIOS:611"
output: html_document
date: "2023-12-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Understanding the Data: 
The North Carolina State Center for Health Statistics (NCSHS) within the North Carolina Department of Health and Human Services collects and reports a multitude of data on topics such as cancer, minority health,and birth defects. NCSHS is also partnered with states nationwide to provide surveillance data 
which includes the Behavioral Risk Factor Surveillance System (BRFSS), Child Health Assessment Monitoring Program (CHAMP), and Pregnancy Risk Assessment Monitoring System (PRAMS). 

#Scope of this Report: 
Annually, as a part of vital statistics reports, NCSHS also reports mortality data that dates back to 1968. Here we will be analyzing mortality data from 2021 which includes the following variables: 

- Place of Death 
- County of Occurrence
- City of Occurrence
- Decedent's Residence County 
- Decedent's Residence City 
- Sex 
- Race of Decedent 
- Age at Death 
- Decedent's Education 
- Place of Injury (if injury death)
- Veteran Status
- Death Month 
- Death Date 
- Death Year

```{r setting up the data, include=FALSE}

setwd("~/work/Desktop/final_project")
install.packages("tidyverse")
library(tidyverse) 
NC_VS_2021 <- read_csv("project_source_data/North_Carolina_Vital_Statistics_2021.csv")

```

## Plot 1: 
The first plot looks at the total counts of mortality in NC in 2021 stratified by sex.
```{r plot 1, include=TRUE}

#Deaths by Sex
p = ggplot(NC_VS_2021, aes(x=SEX)) +
  geom_histogram(stat="count") +
  labs(x = "Sex", y = 'Count') +
  theme(axis.text.x = element_text());

p
```


## Plot 2: 

``` {r plot 2, include=TRUE}
#Death by Age
p = ggplot(NC_VS_2021, aes(x=AGEYRS)) +
  geom_histogram(stat="count") +
  labs(x = "Age", y = 'Count') +
  theme(axis.text.x = element_text());

p


```

## Plot 3: 

``` {r plot 3, include = TRUE}
#Death by Race
p = ggplot(NC_VS_2021, aes(x=RACER)) +
  geom_histogram(stat="count") +
  labs(x = "Race", y = 'Count') +
  theme(axis.text.x = element_text());

p

```

## Plot 4: 

``` {r plot 4, include = TRUE}
#Death by County
p = ggplot(NC_VS_2021, aes(x=DHTMO)) +
  geom_histogram(stat="count") +
  labs(x = "Month", y = 'Count') +
  theme(axis.text.x = element_text());

p

```



## Future Questions